% Appendix — no page limit. Move detail here aggressively.

\section{Full file:line citation table}
\label{app:citations}

\todo{Move the full citation table from
\texttt{GENERATION\_FAILURE\_ANALYSIS.md} \S\emph{Provenance} here as a
LaTeX table. Include the byte-identity note (sampling.py is identical between
\texttt{code\_from\_authors/} and \texttt{fts-diffusion-ref/} modulo one tqdm
import).}

\section{Complete underspecifications list}
\label{app:underspecs}

\todo{Reproduce the 11 entries from \texttt{paper\_underspecifications.md} in
full. Main text \S\ref{sec:repro:underspecs} carries an abridged 6-8-row
version; this appendix carries them all.}

\section{Per-asset detailed metrics}
\label{app:per_asset}

\todo{Tables for GOOG, ZC=F mirroring the S\&P\,500 tables in the main text.}

\section{HPC experimental setup}
\label{app:hpc}

\todo{Document the \texttt{raw\_2/} TATR/TMTR HPC runs: 27 runs each on CSCS
(\texttt{course\_00241}), protocols \emph{author\_independent} and
\emph{continuous\_chunked}. Include the settings dict from
\texttt{task\_NNN\_metadata.json}.}

\section{H2 / H3 / H2 ablations}
\label{app:ablations}

\todo{If experiments E12 (seeds) and/or E13 (capacity ablation) ran, report
results here. If not, state explicitly that they remain future work and
explain why under the deadline.}

\section{Sampling-pipeline design discussion}
\label{app:sampling_design}

\todo{Optional: pull content from \texttt{SAMPLING\_DESIGN\_ANALYSIS.tex}
(teacher-free closed-loop rollout). Position as a candidate \emph{architectural}
fix that future work could explore, distinct from this paper's
\emph{evaluation} contribution.}
