\section{Conclusions}

\begin{frame}{Contributions}
\begin{itemize}\setlength\itemsep{4pt}
  \item Faithful from-scratch replication of FTS-Diffusion (SISC $+$ PGM $+$ PEM) on three assets: S\&P 500, GOOG, ZC$=$F.
  \item Identified four reproducibility issues in the public reference implementation; documented patches.
  \item Defined and compared three synthetic-data protocols (authors / split / random\_init); showed the headline trend depends on the protocol.
  \item Added per-run MAPE confidence intervals and a paper-vs.-replication gap diagnostic.
  \item Pattern dictionaries learned and visualised for $K\!\in\!\{11,13,14\}$ across all three assets, plus MC transition-state figures.
\end{itemize}
\note{Five contributions, all reproducible from the repo. The protocol comparison is the most novel piece of the work: it is not in the paper but it should be.}
\end{frame}

\begin{frame}{Open questions}
\begin{itemize}\setlength\itemsep{4pt}
  \item Which protocol \emph{should} be the benchmark? \texttt{split} is the only one that exploits cross-year MC evolution, but stationarity is not guaranteed.
  \item Is the PEM Markov chain stationary in practice, and on what timescale?
  \item Why does MAPE grow with augmentation under the authors' protocol: leakage from the rolling-window scheme, or genuine distributional drift?
  \item How sensitive are downstream results to $K$? Qualitative flips on ZC$=$F between $K\!=\!11$ and $K\!=\!13$.
\end{itemize}
\begin{exampleblock}{Bottom line}
The architecture is sound; the experimental protocol is not. A clean replication requires both.
\end{exampleblock}
\note{Closing thought: FTS-Diffusion has the right inductive bias (motifs as variable-length DTW clusters) but the published downstream comparison is fragile under protocol choice.}
\end{frame}

\begin{frame}[fragile]{Backup: repository layout}
\footnotesize
\begin{verbatim}
ml-in-fcs/
  architectures/{asset}/k{K}/{data,res,trained_models}/   <- SISC + PGM + PEM ckpts
  synthetic/{asset}/k{K}/{protocol}/run_XX_{syn,states}.npy
  results/tatr/{asset}/k{K}/{protocol}/run_XX.csv          <- per-run MAPE
  figures/{sisc,tatr,mc_analysis}/{asset}/k{K}/...         <- final PDFs
  src/{TATR_Reduced_Replication.ipynb,
       FTS_Diffusion_Training.ipynb,
       B3_SISC_Simulated_Investigation.ipynb}
\end{verbatim}
\end{frame}

\begin{closingframe}

Deli Lin, Eric Shao, Lapo Linossi, Tom Haidinger\\
\medskip
ETH Zurich\\
Machine Learning in Finance and Complex Systems\\
Supervisor: Alexander Arandjelovic

\medskip

Thank you. \emph{Questions?}

\end{closingframe}
